Article Title

### 2013-12-30 plos.tex
new version of the discussion nearly completed.

\DeclareMathOperator{\Tr}{tr}
\newcommand{\mcond}{\,\middle\vert\,}
\newcommand{\cond}{\,\vert\,}
\newcommand{\figref}[2]{Fig.\;\ref{fig:#1}\,#2}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)} 
\newcommand{\eloss}[1]{\mathcal L_0\left(#1\right)}
\newcommand{\T}{{\sf T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\newcommand{\TODO}[1]{\emph{\small\color{blue}$\langle\langle$#1$\rangle\rangle$}}
\newcommand*\dif{\mathop{}\,d}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\rank}{rank}

@@

In this study, we sought most statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes of about 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix would be most efficiently represented by a sparse network of partial correlations between pairs of observed neurons combined with common latent factors representing unobserved inputs and global network fluctuations.  Indeed, in cross-validation tests, the covariance matrix estimator based on such approximations outperformed other regularized estimators. Mirroring previously described patterns of synaptic connectivity, the density of positive partial correlations decreased rapidly with both physical distances and preferred orientation differences whereas negative partial correlations were less selective.  Because of its superior estimation performance, the inferred partial correlation structure obtained using these methods may provide a better description of the functional connectivity in dense circuits than the usual sample correlations. 

@@

Pearson correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are among the most familiar descriptive statistics of neural population activity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  For example, \emph{noise correlations}, \emph{i.e.}~the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect key features of functional connectivity in neural circuits and reflect interactions between neurons \cite{Gerstein:1964}.  This interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distances between neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010,Ko:2011, Smith:2013b}, cortical layer specificity \cite{Hansen:2012,Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011}, changes due to sensory stimulation and global brain states \cite{Goard:2009, Kohn:2009, Ecker:2010, Renart:2010}, and others.

@@

For example, one form of low-dimensional representation of the correlation matrix is its low-rank approximation. Such approximations are particularly suitable for capturing shared fluctuations across the recorded population, which may arise in local circuits or be externally driven. Low-rank components can be extracted using principal component or factor analysis.

Alternatively, the correlation matrix can be expressed through the partial pairwise correlations. The partial correlation between a pair of neurons is a measure of their linear association after the linear effects of all other neurons have been taken into account. Partial correlations carry particular significance when the relationship between the variables is approximately linear. In this case, small pairwise partial correlations indicate that the two variables are nearly independent when conditioned on the activity of the other neurons.

Conditional independence between a pair of variables suggests a lack of direct interaction between the underlying processes. Therefore, networks of partial correlations (sometimes called \emph{association networks}) can be used to infer interactions between the components of a complex system.  This approach has been used to uncover the structure of gene interaction networks \cite{Schafer:2005,Peng:2009} and the functional connectivity between brain regions from fMRI signals \cite{Varoquaux:2012,Ryali:2012}. In multivariate normal distributions (or, more generally, \emph{elliptical distributions}) interactions are completely defined by linear effects and, therefore, partial correlations reflect conditional dependencies.  A multivariate normal distribution specified by a weighted graph of non-zero partial correlations (along with the means and variances of individual neurons) is known as a Gaussian Graphical Model or Gauss-Markov Random Field \cite{Koller:2009}. With departure from linearity, the correspondence between conditional dependence and partial correlations diminishes and can break down \cite{Loh:2012}. For example, the coupling terms in pairwise Ising models do not directly correspond o the partial correlations \cite{Schneidman:2006, Tkacik:2006}.

@@

Given a covariance matrix estimate $C$ (not necessarily the sample covariance $C_{\sf 0}$), the correlation matrix $R$ is calculated by normalizing $C$ by its diagonal (variance estimates):
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2}
\end{equation}
Similarly, the matrix of partial correlations $P$ is computed as the negative normalized version of the \emph{precision matrix} (inverse covariance matrix) $C^{-1}$:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}
Here $\circ$ denotes entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix. Clearly, off-diagonal zeros in the precision matrix correspond to zero partial correlations.

@@

Estimation can be improved through \emph{regularization}: deliberately imposing some kind of structure on the estimate. The sample estimate $C_{\sf 0}$, which is usually used to estimated covariance matrices in neuroscience, is unbiased ($\E{C_0}=\Sigma$).  However, on average, it falls far from $\Sigma$ due to its sensitivity to sampling noise.  Low-dimensional estimates are typically less susceptible to sampling noise but can introduce biases. Regularization strikes a favorable balance between bias and variability

 A regularized estimate is produced by biasing (`shrinking’) the estimate from the unbiased estimate toward a selected low-dimensional approximation (`target’) of the unbiased estimate \cite{Schafer:2005, Bickel:2006}.  The selection of the optimal target estimate and the optimal amount of shrinkage can be optimized based on the training data by analytical methods or by cross-validation.  However, some regularization schemes only perform target selection (\emph{dimensionality reduction}) and replace the estimate without shrinkages whereas others only perform shrinkage given a single target estimate (\emph{shrinkage estimators}).

@@

\paragraph{Illustration of estimation of a neural correlation matrix}
To illustrate the challenges and techniques of estimation of the correlation matrix from a finite sample, we consider a regularization scheme based on \emph{covariance selection} (Figure \ref{fig:01}) \cite{Dempster:1972}. Here the estimate is produced by fitting only an optimal subset of the coefficients of the precision matrix while setting the rest to zero. We estimated the covariance matrix of somatic calcium signals of a group of 298 neurons in mouse visual cortex using both the sample covariance matrix (\figref{01}{A, B}) and covariance selection (\figref{01}{C}). Due to the low-pass filtration effect of calcium dye kinetics, correlations in unprocessed calcium signals are higher than typical firing rate correlations on shorter temporal scales. The close similarity of the two estimates of the correlation matrix (\figref{01}{B and C}) belies the utterly different partial correlation structure (\figref{01}{D and E}): The regularized estimate is produced from the same data by setting to zero 31,501 (71.2\%) of the off-diagonal coefficients of the precision matrix and fitting the remaining 12,752 coefficients. Estimation by covariance selection is attractive for several reasons: First, with fewer free parameters it is less susceptible to sampling noise, yielding estimates that are, on average, closer to the truth. Additionally, if we assume that partial correlations do indeed reflect elements of the functional connectivity in the circuit, the second estimate also appears more plausible: in neural circuits, each neuron only interacts with a fraction of the entire population.

\begin{figure}[!ht]    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Illustration of regularized estimation of partial correlations.}
        {\bf A}. The sample correlation matrix of unprocessed somatic calcium signals from a population of cells in mouse visual cortex.
        The outlined square fragment is magnified in {\bf B}.
        {\bf C}. The same fragment of another estimate of the correlation matrix regularized to yield sparse partial correlations.
        Corresponding fragments of partial correlations matrices of the unregularized and regularized estimated are shown in {\bf D} and {\bf E}, respectively.
    }
    \label{fig:01}}
    {\includegraphics[width=8.3cm]{./figures/Figure01.pdf}}
\end{figure}
\paragraph{Our contribution}
Various other regularization schemes also produce improved estimates.  In this study, we compared the sample covariance matrix and four regularized estimators based on distinct correlation structures as follows:
\begin{description}
\item[$C_{\sf 0}$] -- sample covariance matrix, the usual estimator used in previous studies.
\item[$C_{\sf diag}$] -- linear shrinkage toward a diagonal covariance matrix.
\item[$C_{\sf factor}$] -- linear shrinkage toward a low-rank matrix produced by factor analysis with an optimal number of factors.
\item[$C_{\sf sparse}$] -- sparse inverse \emph{covariance selection} \cite{Dempster:1972, Friedman:2008}, without shrinkage.   
\item[$C_{\sf sparse+latent}$] -- inverse matrix as the sum of a sparse  and a low-rank component \cite{Chandrasekaran:2010, Ma:2013}.
\end{description}

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Estimators whose low-dimensional regularization targets can represent the structure of the true covariance matrix outperform other estimators.}
        {\bf Row 1.} Graphical representations of four types of low-dimensional structures of interactions between observed neurons (green spheres) and latent units (light-shaded spheres).
        In the `independent model' ({\bf A}), observed neurons exert no linear effects on one another neither directly nor through interactions with common latent units. 
        In `latent factors' ({\bf B}), the correlated activity of observed cells is driven by several latent units. 
        In `sparse interactions' ({\bf C}), the correlation matrix is defined by a set of linear interactions between observed neurons. 
        In `sparse+latent' ({\bf D}), correlations arise through direct linear interactions in a sparse subset of observed neurons and through interactions with common latent units. 
        {\bf Row 2.} Examples of $50\times 50$ correlation matrices corresponding to each type of low-dimensional structure. 
        The factor model ({\bf B}) has three latent units. 
        The partial correlation matrix of the sparse model ({\bf C}) is 73\% sparse.
        The `sparse+latent' matrix has one latent unit and its direct interactions are 78\% sparse.
        {\bf Row 3.} Examples of sample correlation matrices calculated from samples of 1000 observations taken from simulated random processes with corresponding correlation matrices from row 2.
        {\bf Row 4.} Mean \emph{excess loss} (Eq.~\ref{eq:excess-loss}) attained by each of the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.
        {\bf Row 5.} Mean \emph{cross-validation loss} (Eq.~\ref{eq:rel-cv-loss}) of covariance estimators with respect to the matching estimator. The values are relative to the validation loss of the estimator that matches the low-dimensional structure of the true covariance matrix. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

@@

With empirical data, acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation is used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as a training dataset. The validation losses from each of such `folds' are averaged to produce \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In the present formulation, $\ell_C$ is known as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier).

@@

In simulation, CV-loss accurately reproduced the differences between the estimators' excess losses, although with greater variability (Figure \ref{fig:02}, Row 5). For each kind of ground truth (Row 2), relative CV-losses were computed with respect to the estimator whose regularization target matched the structure of the respective ground truth. Just as with excess loss in Row 4, the means and their standard errors were computed from 30 samples taken for each ground truth and for each sample size.

@@

\begin{figure}[!ht]    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf The sparse+latent estimator $C_{\sf sparse+latent}$ outperforms the other estimators on neural data.}
    {\bf A--D.} Histograms of average cross-validation loss differences of the respective estimators $C_{\sf 0}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ from $C_{\sf sparse+latent}$. 
    The histograms are based on 31 imaged sites in 24 mice. 
    All medians (red dashed lines) were significantly greater than zero, indicating the dominance of $C_{\sf sparse+latent}$ over the other estimators. 
    The arrow heads indicate the results for the site shown in Fig.~\ref{fig:03} and Fig.~\ref{fig:05}
{\bf A} $C_{\sf sparse+latent}$ outperforms $C_{\sf 0}$: median improvement 0.16 nats/neuron, $p=3.9\times 10^{-5}$.
    {\bf B} $C_{\sf sparse+latent}$ outperforms $C_{\sf diag}$: median improvement 0.0032 nats/neuron, $p=2.2\times 10^{-3}$.
    {\bf C} $C_{\sf sparse+latent}$ outperforms $C_{\sf factor}$: median improvement 0.16 nats/neuron, $p=6.9\times 10^{-5}$.
    {\bf D} $C_{\sf sparse+latent}$ outperforms $C_{\sf sparse}$: median improvement 0.0016 nats/neuron, $p=5.5\times 10^{-6}$.
} \label{fig:04}}
    {\includegraphics{./figures/Figure04.pdf}}
\end{figure}

@@

We examined the structure of the sparse+latent estimate at individual sites (an example site is depicted in Fig.\;\ref{fig:03}). At these sites the regularized estimate of the correlation matrix appeared very similar to the unregularized sample correlation matrix (\figref{05}{A and D}). However, the corresponding partial correlation matrices differed dramatically (\figref{05}{B and E}). The partial correlation was decomposed into sparse and low-rank components (\figref{05}{C}). Although correlations were mostly positive, the sparse partial correlations (or `interactions'), had a much larger fraction of negative values than sample correlations. The sparse component had 82.1\% sparsity (or, conversely, 17.9\% connectivity), which corresponded to the average node degree (interactions per cell) of 52.5 (\figref{05}{G}). The low-rank component had rank 17.

@@

\section*{Discussion}
\paragraph{Functional and anatomical connectivity}
Systems neuroscience is the quest to discover principles whereby anatomical organization of neural circuits gives rise to their function.  This line of inquiry requires expressive descriptions of the observed activity of neural circuits.

The term \emph{functional architecture} most commonly refers to stimulus response properties of neurons \cite{Reid:2012}. Over the past 50 years, tremendous progress has been made in relating the functional architecture to the anatomical, synaptic wiring of neuronal populations and individual neurons.  This line of inquiry is based on the view of the brain as a complex filter of external stimuli.

In contrast to functional architecture, \emph{functional connectivity} commonly refers to the degree of correlation, synchronization, or other forms of statistical association between neuronal populations or individual neurons inferred from spontaneous activity or trial-to-trial variations,  not directly driven by external stimuli.  Functional connectivity is thought to reflect internal, recurrent connectivity in neural circuits.   However, less experimental evidence has been garnered to for links between functional and anatomical connectivity.

Neural correlations boast a rich history as indicators of functional connectivity. When measured on temporal scales of milliseconds, sharp peaks or troughs in the cross-correlogram between pairs of neurons at latencies typical of synaptic transmission provide evidence of monosynaptic connection between pairs of cells \cite{Moore:1970, Alonso:1998, Denman:2013} (although other processes can also generate sharp peaks in the cross-correlogram while many monosynaptic connections may not produce such signatures). Without targeted stimulation of specific cells or measurements of subthreshold intracellular events, such analysis remains the best method for detecting probable monosynaptic connections in vivo.

When measured at longer temporal scales, neural correlations reflect activations involving large ensembles of neurons and may reflect direct causal anatomical interactions, indirect causal interactions, or complex network-wide emergent features of activity with little correspondence to the local anatomical connectivity. Calcium imaging allows simultaneous measurements of the activity of large populations of cells \emph{in vivo} \cite{Katona:2012, Cotton:2013} and even of entire nervous systems \cite{Leung:2013, Ahrens:2013}, but its temporal bandwidth is limited to the scale of tens or hundreds milliseconds by the kinetics of calcium-sensitive dyes, limiting its use to slower temporal scales at which correspondence between functional and anatomical connectivity is less direct.

>>>>

\DeclareMathOperator{\Tr}{tr}
\newcommand{\sq}[1]{\lq#1\rq}
\newcommand{\mcond}{\,\middle\vert\,}
\newcommand{\cond}{\,\vert\,}
\newcommand{\figref}[2]{Fig.\;\ref{fig:#1}\,#2}
\newcommand{\loss}[1]{\mathcal L\left(#1\right)} 
\newcommand{\eloss}[1]{\mathcal L_0\left(#1\right)}
\newcommand{\T}{{\sf T}}
\newcommand{\E}[2][]{\mathbb E_{#1}\left[ #2\right]}    % expected value
\newcommand{\TODO}[1]{\emph{\small\color{blue}$\langle\langle$#1$\rangle\rangle$}}
\newcommand*\dif{\mathop{}\,d}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator{\rank}{rank}

@@

In this study, we sought most statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes of about 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix would be most efficiently represented by a sparse network of partial correlations between pairs of observed neurons combined with common latent factors representing unobserved inputs and global network fluctuations.  Indeed, in cross-validation tests, the covariance matrix estimator based on such approximations outperformed other regularized estimators. Mirroring previously described patterns of synaptic connectivity, the density of positive partial correlations decreased rapidly with both physical distances and preferred orientation differences whereas negative partial correlations were less selective.  Because of its superior estimation performance, the partial correlation structure by these methods will likely provide a better description of the functional connectivity in dense circuits than the usual sample correlations. 

@@

Pearson correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are among the most familiar descriptive statistics of neural population activity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  For example, \emph{noise correlations}, \emph{i.e.}~the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect key features of functional connectivity in neural circuits and reflect interactions between neurons \cite{Gerstein:1964}.  This interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distances between neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010,Ko:2011, Smith:2013b}, cortical layer specificity \cite{Hansen:2012,Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011}, changes due to sensory stimulation and global brain states \cite{Goard:2009, Kohn:2009, Ecker:2010, Renart:2010}, and others.

@@

For example, one form of low-dimensional representation of the correlation matrix is its low-rank approximation. Such approximations are particularly suitable for capturing shared fluctuations across the recorded population, which may arise in local circuits or be externally driven. Low-rank components can be extracted using principal component or factor analysis \cite{Chapin:1999, Peyrache:2010, Lopes:2011, Lopes:2013}.

Alternatively, the correlation matrix can be expressed through the partial pairwise correlations. The partial correlation between a pair of neurons is a measure of their linear association after the linear effects of all other neurons have been taken into account. Partial correlations carry particular significance when the relationship between the variables is approximately linear. In this case, a low pairwise partial correlation indicates that the two variables are nearly independent when conditioned on the activity of the other neurons.

Conditional independence between a pair of variables suggests a lack of direct interaction between the underlying processes. Therefore, networks of partial correlations (sometimes called \emph{association networks}) can be used to infer interactions between the components of a complex system.  This approach has been used to uncover the structure of gene interaction networks \cite{Schafer:2005, Peng:2009} and the functional connectivity between brain regions from fMRI signals \cite{Varoquaux:2012, Ryali:2012}. In multivariate normal distributions (or, more generally, \emph{elliptical distributions}) interactions are completely defined by linear effects and, therefore, partial correlations reflect conditional dependencies.  A multivariate normal distribution specified by a weighted graph of non-zero partial correlations (along with the means and variances of individual neurons) is known as a Gaussian Graphical Model or Gauss-Markov Random Field \cite{Koller:2009}. With departure from linearity, the correspondence between conditional dependence and partial correlations diminishes and can break down \cite{Loh:2012}. For example, the coupling terms in pairwise Ising models fitted to the same sample correlation matrix will produce a different network of dependencies than that inferred by the partial correlations \cite{Schneidman:2006, Tkacik:2006}.

@@

Given a covariance matrix estimate $C$ (not necessarily the sample covariance $C_{\sf 0}$), the correlation matrix $R$ is calculated by normalizing $C$ by its diagonal (variance estimates):
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2}
\end{equation}
Similarly, the matrix of partial correlations $P$ is computed by normalizing the the negative \emph{precision matrix} $C^{-1}$ (inverse of the covariance matrix):
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}
Here $\circ$ denotes entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix. Clearly, off-diagonal zeros in the precision matrix correspond to zero partial correlations.

@@

Estimation can be improved through \emph{regularization}: deliberately imposing some kind of structure on the estimate. The sample estimate $C_{\sf 0}$, which is usually used to estimate covariance matrices in neuroscience, is unbiased ($\E{C_0}=\Sigma$).  However, on average, it falls far from $\Sigma$ due to its sensitivity to sampling noise.  Low-dimensional estimates are typically less susceptible to sampling noise but can introduce biases. Regularization strikes a favorable balance between bias and variability

 A regularized estimate is produced by \sq{shrinking} the unbiased estimate toward a selected low-dimensional approximation (\sq{target estimate}) of the unbiased estimate \cite{Schafer:2005, Bickel:2006}.  The selection of the optimal target estimate and the optimal amount of shrinkage can be deduced from the training data analytically or by cross-validation.  Some regularization schemes only perform target selection (\emph{dimensionality reduction, variable selection, or feature selection}) and replace the estimate completely, without shrinkage. Others only perform shrinkage toward a single target estimate (\emph{shrinkage estimators}).

@@

Although regularized covariance estimates have become established in other fields such as finance \cite{Ledoit:2003}, functional genomics \cite{Schafer:2005}, and brain imaging \cite{Ryali:2012}, surprisingly little work has been done to identity optimal regularization schemes for neural correlations.  Since the performance of regularization schemes depends on the dominant regularities in the data, their optimal choice depends on the type of neural recording, the temporal scale, and brain states. For example, dense recordings from local circuits with high temporal resolution in desynchronized brain states may warrant greater emphasis on partial correlations, reflecting direct interactions between neurons.  Recordings from distributed groups of cells, synchronous circuit activity, or recordings with low temporal resolution are likely to reveal common correlated activity across many cells with correlation matrices best modeled by low-rank components.  Therefore, the optimal choice of the regularization scheme, ascertained from the data by cross-validation, constitutes a statistical description of a neural circuit that informs about the dominant regularities in the data.
\paragraph{Illustration of estimation of a neural correlation matrix}
To illustrate the challenges and techniques of estimation of the correlation matrix from a finite sample, we consider a regularization scheme based on \emph{covariance selection} (Figure \ref{fig:01}) \cite{Dempster:1972}. Here the estimate is produced by fitting only an optimal subset of the coefficients of the precision matrix while setting the rest to zero. We estimated the covariance matrix of somatic calcium signals of a group of 298 neurons in mouse visual cortex using both the sample covariance matrix (\figref{01}{A, B}) and covariance selection (\figref{01}{C}). Due to the low-pass filtration effect of calcium dye kinetics, correlations in unprocessed calcium signals are higher than typical firing rate correlations on shorter temporal scales. The close similarity of the two estimates of the correlation matrix (\figref{01}{B and C}) belies the utterly different partial correlation structure (\figref{01}{D and E}): The regularized estimate is produced from the same data by setting to zero 31,501 (71.2\%) of the off-diagonal coefficients of the precision matrix and fitting the remaining 12,752 coefficients. Estimation by covariance selection is attractive for several reasons: First, with fewer free parameters it is less susceptible to sampling noise, yielding estimates that are, on average, closer to the truth. Additionally, if we assume that partial correlations do indeed reflect elements of the functional connectivity in the circuit, the second estimate also appears more plausible: in neural circuits, each neuron only interacts with a fraction of the entire population.

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Illustration of regularized estimation of partial correlations.}
        {\bf A}. The sample correlation matrix of unprocessed somatic calcium signals from a population of cells in mouse visual cortex.
        The outlined square fragment is magnified in {\bf B}.
        {\bf C}. The same fragment of another estimate of the correlation matrix regularized to yield sparse partial correlations.
        Corresponding fragments of partial correlations matrices of the unregularized and regularized estimated are shown in {\bf D} and {\bf E}, respectively.
    }
    \label{fig:01}}
    {\includegraphics[width=8.3cm]{./figures/Figure01.pdf}}
\end{figure}
\paragraph{Our contribution}
Various other regularization schemes also produce improved estimates.  In this study, we compared the sample covariance matrix and four regularized estimators based on distinct correlation structures as follows:
\begin{description}
\item[$C_{\sf 0}$] -- sample covariance matrix, the usual estimator used in previous studies.
\item[$C_{\sf diag}$] -- linear shrinkage toward a diagonal covariance matrix.
\item[$C_{\sf factor}$] -- linear shrinkage toward a low-rank matrix produced by factor analysis with an optimal number of factors.
\item[$C_{\sf sparse}$] -- sparsified precision matrix,  \emph{covariance selection} \cite{Dempster:1972, Friedman:2008}, without shrinkage.   
\item[$C_{\sf sparse+latent}$] -- precision matrix composed as the sum of a sparse and a low-rank component \cite{Chandrasekaran:2010, Ma:2013}.
\end{description}

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Estimators whose low-dimensional regularization targets can represent the structure of the true covariance matrix outperform other estimators.}
        {\bf Row 1.} Graphical representations of four types of low-dimensional structures of interactions between observed neurons (green spheres) and latent units (light-shaded spheres).
        In the \sq{independent model} ({\bf A}), observed neurons exert no linear effects on one another neither directly nor through interactions with common latent units. 
        In \sq{latent factors} ({\bf B}), the correlated activity of observed cells is driven by several latent units. 
        In \sq{sparse interactions} ({\bf C}), the correlation matrix is defined by a set of linear interactions between observed neurons. 
        In \sq{sparse+latent} ({\bf D}), correlations arise through direct linear interactions in a sparse subset of observed neurons and through interactions with common latent units. 
        {\bf Row 2.} Examples of $50\times 50$ correlation matrices corresponding to each type of low-dimensional structure. 
        The factor model ({\bf B}) has three latent units. 
        The partial correlation matrix of the sparse model ({\bf C}) is 73\% sparse.
        The \sq{sparse+latent} matrix has one latent unit and its direct interactions are 78\% sparse.
        {\bf Row 3.} Examples of sample correlation matrices calculated from samples of 1000 observations taken from simulated random processes with corresponding correlation matrices from row 2.
        {\bf Row 4.} Mean \emph{excess loss} (Eq.~\ref{eq:excess-loss}) attained by each of the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.
        {\bf Row 5.} Mean \emph{cross-validation loss} (Eq.~\ref{eq:rel-cv-loss}) of covariance estimators with respect to the matching estimator. The values are relative to the validation loss of the estimator that matches the low-dimensional structure of the true covariance matrix. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

@@

With empirical data, acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation is used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as a training dataset. The validation losses from each of such \sq{folds} are averaged to produce \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In the present formulation, $\ell_C$ is known as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier).

@@

In simulation, CV-loss accurately reproduced the differences between the estimators' excess losses, although with greater variability (Figure \ref{fig:02}, Row 5). For each kind of ground truth (Row 2), relative CV-losses were computed with respect to the estimator whose regularization target matched the structure of the respective ground truth.

@@

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf The sparse+latent estimator $C_{\sf sparse+latent}$ outperforms the other estimators on neural data.}
    {\bf A--D.} Histograms of average cross-validation loss differences of the respective estimators $C_{\sf 0}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ from $C_{\sf sparse+latent}$. 
    The histograms are based on 31 imaged sites in 24 mice. 
    All medians (red dashed lines) were significantly greater than zero, indicating the dominance of $C_{\sf sparse+latent}$ over the other estimators. 
    The arrow heads indicate the results for the site shown in Fig.~\ref{fig:03} and Fig.~\ref{fig:05}
{\bf A} $C_{\sf sparse+latent}$ outperforms $C_{\sf 0}$: median improvement 0.16 nats/neuron, $p=3.9\times 10^{-5}$.
    {\bf B} $C_{\sf sparse+latent}$ outperforms $C_{\sf diag}$: median improvement 0.0032 nats/neuron, $p=2.2\times 10^{-3}$.
    {\bf C} $C_{\sf sparse+latent}$ outperforms $C_{\sf factor}$: median improvement 0.16 nats/neuron, $p=6.9\times 10^{-5}$.
    {\bf D} $C_{\sf sparse+latent}$ outperforms $C_{\sf sparse}$: median improvement 0.0016 nats/neuron, $p=5.5\times 10^{-6}$.
} \label{fig:04}}
    {\includegraphics{./figures/Figure04.pdf}}
\end{figure}

@@

We examined the structure of the sparse+latent estimate at individual sites (an example site is depicted in Fig.\;\ref{fig:03}). At these sites the regularized estimate of the correlation matrix appeared very similar to the unregularized sample correlation matrix (\figref{05}{A and D}). However, the corresponding partial correlation matrices differed dramatically (\figref{05}{B and E}). The partial correlation was decomposed into sparse and low-rank components (\figref{05}{C}). Although correlations were mostly positive, the sparse partial correlations (or \sq{interactions}), had a much larger fraction of negative values than sample correlations. The sparse component had 82.1\% sparsity (or, conversely, 17.9\% connectivity), which corresponded to the average node degree (interactions per cell) of 52.5 (\figref{05}{G}). The low-rank component had rank 17.

@@

\section*{Discussion}
\paragraph{Functional connectivity}
How the anatomical organization of neural circuits gives rise to their function is a foremost question in neuroscience.  Effective phenomenological descriptions of neural activity, once discovered, enable investigation of underlying physiological mechanisms.  For example, the discoveries of receptive fields and tuning properties of individual cells, over the past 60 years, have led to fruitful hypotheses about the underlying patterns of synaptic connectivity, principles of neural coding, and to the discovery of other organizational principles such as various cortical feature maps \cite{Reid:2012}.

The term \emph{functional connectivity} encompasses various phenomenological descriptions of patterns of activity in populations of neurons.  It is commonly expressed as the degree of correlation, synchronization, or other forms of statistical association between neurons that is not accounted by the independent response properties of the individual cells.  Neuroscientists have hypothesized that spontaneous activity and trial-to-trial variations in population responses are not random noise but endogenously generated activity with fundamental roles in stimulus coding, sensory inference, attention, learning, and cognition and that further progress depends on understanding of patterns of activity in populations of cells.  This line of inquiry may require characterizing changes in functional connectivity due to development, learning, brain states, and behavioral tasks and finding its relationships to the circuits’ anatomical organization. 
\paragraph{Temporal scales of functional connectivity}
Analyses of functional connectivity depend profoundly on their temporal resolution.  At millisecond-scale resolution, cross-correlograms of the spiking activity of pairs of cells manifest statistical signatures of monosynaptic connections \cite{Gerstein:1964, Moore:1970, Gerstein:1989, Alonso:1998, Denman:2013}.  Other methods require temporal resolution on the order of membrane time constants (milliseconds to tens of milliseconds): \emph{e.g.}~statistical models that mimic membrane integration (\cite{Pillow:2008}) and information-theoretic and statistical physics models that bin spike trains finely to ensure a binary representation \cite{Schneidman:2006, Tkacik:2006, Tkacik:2013}. At slower scales of tens to hundreds milliseconds, temporal sequencing of spikes can no longer resolve directions of monosynaptic interactions and any expression of functional connectivity summarizes chains of synaptic transmissions with more complex, emergent properties of network computation: the analysis shifts from to detecting co-active cell assemblies and the low-rank structure of correlations \cite{Ch:2010, Lopes:2011, Lopes:2013}.
